Phylodynamic inference of contact network parameters through approximate Bayesian computation

### 2016-06-07 doc/thesis/intro.tex

{\color{blue}\uline{
Recall that the primary objective of our work is to develop a statistical
inference method for estimating contact network parameters from transmission
trees. In particular, for a network model with parameters $\theta$ and an input
transmission tree $T$, we are interested in the \emph{posterior distribution}}
\begin{align}
  \Pr(\theta \mid T) = \frac{\Pr(T \mid \theta) \Pr(\theta)}{\Pr(T)}.
  \label{eq:post}
\end{align}
\uline{As outlined in} \cref{sec:obj}, \uline{both the likelihood $\Pr(T \mid
\theta)$ and the normalizing constant $\Pr(T)$ are likely computationally
intractable. Hence, rather than computing the posterior distribution
analytically, we will approximate it using a \defn{Monte Carlo} approach.
The fundamental idea behind Monte Carlo methods is succinctly expressed by
\textcite{liu2001theoretical}}:
\begin{quote}
    \uline{Monte Carlo's view of the world is that any probability distribution $\pi$, 
    regardless of its complexity, can always be \emph{represented} by a
    discrete sample from it. By ``represented'', we mean that any computation
    of expectations using $\pi$ can be replaced to an acceptable degree of
    accuracy by using the empirical distribution resulting from the discrete
    sample.}
\end{quote}
\uline{In other words, if we are able to sample enough points from a
distribution of interest, we will be able to make reasonably accurate
statements about the distribution itself. For example, the expected value of
the distribution can be estimated by the sample's population mean. The reason
Monte Carlo methods will be useful in this work is that algorithms exist for
obtaining samples from distributions, including as the aforementioned
posterior, that are analytically intractable and from which direct sampling is
not possible } (for a review see \autocite{robert2004monte}).
\uline{\Gls{SMC}}~\autocite{doucet2000sequential, doucet2001introduction,
liu2008monte} \uline{is one such algorithm.}

@@

\uline{In this work, the distribution of interest is the posterior distribution}
\ref{eq:post}. \uline{The particles are particular values of the parameters
$\theta$ of the contact network model being studied. If we were taking a
typical Bayesian Monte Carlo approach to this problem, the particles would end
up weighted by their posterior probability and distributed in such a way that
the weighted population was a reasonable representation of $\Pr(\theta \mid
T)$. This turns out to be not straightforward in our case, as we will address
in} \cref{sec:abc}. \uline{For now, nothing is lost by \ldots}}

{\color{blue}\uline{
The goal of this section of the introduction is to describe an algorithm called
the \gls{SMC} sampler~\autocite{del2006sequential}, which forms the basis of
the adaptive \gls{ABC}-\gls{SMC} algorithm we apply toward the main objective
of this work. We begin by describing \gls{SIS}, which is a precursor to
\gls{SMC} that samples from a sequence of distributions defined on spaces of
increasing dimension. We then describe \gls{SMC} itself, which extends
\gls{SIS} with a resampling step to fight particle degeneracy. Finally, we
outline the \gls{SMC} sampler, which allows \gls{SMC} to be applied to
sequences of distributions all defined on the same space. This terminology will
become clear as the methods are described.
}}

\Gls{SMC} is the name for a family of statistical inference methods that rely
on approximating probability distributions of interest with large collections
of \defn{particles}, here denoted
$\set{x^{(k)}}$~\autocite{doucet2000sequential, doucet2001introduction,
liu2008monte}. These collections or \defn{populations} are constructed to form
a \defn{Monte Carlo approximation} to some distribution of interest $\pi$,
meaning that the empirical distribution of the particles converges in
distribution to $\pi$ as the population size gets
large~\autocite{liu2001theoretical}. The word \defn{sequential} is used because
the particle populations are modified in an iterative fashion over time, for
example, to incorporate new evidence.

@@

Both algorithms involve progression through a sequence of related
distributions. For \gls{SIS}, these distributions are lower-dimensional
marginals of the target distribution, while for the \gls{SMC} sampler, they are
of the same dimension and constitute a smooth progression from an initial to a
final distribution. In both cases, the neighbouring distributions in the
sequence are related to each other in some way, and we can take advantage of
that relationship to create a sequence of importance distributions alongside
the sequence of targets. In \gls{SIS}, the neighbouring marginals
$\pi(\vec{x_i})$ and $\pi(\vec{x_{i+1}})$ were related by the conditional
density $\pi(x_i \mid \vec{x_{i-1}})$, which we used to inform the importance
distribution. In \gls{SMC}, the relationship between subsequent distributions
is less explicit, but it is assumed that they are related closely enough that
an importance distribution for $\pi_i$ can be easily transformed into one for
$\pi_{i+1}$. In particular, the sequence of importance distributions $\eta_i$
is constructed as
\begin{align}
  \label{eq:impint}
  \eta_i(x') = \int \eta_{i-1}(x) K_i(x, x') \d x,
\end{align}
where $K_i$ is a Markov kernel and the integral is over the space on which the
$\pi_i$ are defined. The choice of $K_i$ should be based on the perceived
relationship between $\pi_{i-1}$ and $\pi_i$. \textcite{del2006sequential}
propose the use of a \gls{MCMC} kernel with equilibrium distribution $\pi_i$.
That is,
\[
  K_i(x, x') = \max\left(1, \frac{q(x', x)\pi_i(x)}{q(x, x')\pi_i(x')}\right),
\]
where $q(x, x')$ is a proposal function such as a Gaussian distribution
centred at $x$ {\color{blue}\uline{from which $x'$ is drawn}} (see
\cref{subsec:mfit}).

Although this method of building up $\eta$ appears straightforward, the
drawback is that the importance distribution itself becomes intractable. In
particular, evaluating $\eta_i(x)$ involves a $i$-dimensional integral of the
type in \cref{eq:impint}. As it is necessary to evaluate $\eta(x)$ pointwise to
perform \gls{IS}, this construction appears to have defeated the purpose of
providing an importance distribution for each $\pi_i$.
\textcite{del2006sequential} overcome this problem with two ``artificial''
objects. First, they propose the existence of \textit{backward} Markov kernels
$L_{i-1}(x_i, x_{i-1})$. For now, these kernels are arbitrary; they will later
be precisely defined on a problem-specific basis. Second, the authors define an
alternative sequence of target distributions
\[
  \tilde{\pi}_i(\vec{x_i}) = \pi_i(x_i) \prod_{k=1}^{i-1} L_k(x_{k+1}, x_k)
\]
of increasing dimension. This brings us back to the setting described above in
\cref{subsec:sis}, namely of building up an importance distribution of
dimension $d$ sequentially through lower-dimensional distributions. We can
write $\tilde{\pi}_i$ in terms of $\tilde{\pi}_{i-1}$ by noticing that
\begin{align*}
  \frac{\tilde{\pi}_i(\vec{x_i})}{\tilde{\pi}_{i-1}(\vec{x_{i-1}})} 
  = \frac{\pi_i(x_i) \prod_{k=1}^{i-1} L(x_{k+1}, x_k)}
  {\pi_{i-1}(x_{i-1}) \prod_{k=1}^{i-2} L(x_{k+1}, x_k)}
  = \frac{\pi_i(x_i) L(x_i, x_{i-1})}{\pi_{i-1}(x_{i-1})},
\end{align*}
and hence
\[
  \tilde{\pi}_i = \frac{\pi_i(x_i) L(x_i, x_{i-1})}{\pi_{i-1}(x_{i-1})} \cdot \tilde{\pi}_{i-1}.
\]
Therefore, the importance weights for these new targets are defined recursively as
\begin{align}
  w(\vec{x_i}) 
    &= \frac{\tilde{\pi}_i(\vec{x_i})}{\eta_i(\vec{x_i})} \\
    &= \frac{\tilde{\pi}_{i-1}(\vec{x_{i-1}}) \pi_i(x_i) L(x_i, x_{i-1})}
           {\eta_{i-1}(\vec{x_{i-1}}) \pi_{i-1}(x_{i-1}) K_i(x_{i-1}, x_i)} \\
    &= w(\vec{x_{i-1}}) \cdot
      \frac{\pi_i(x_i) L_{i-1}(x_i, x_{i-1})}
           {\pi_{i-1}(x_{i-1}) K_i(x_{i-1}, x_i)} \\
    &\propto w(\vec{x_{i-1}}) \cdot
      \frac{\gamma_i(x_i) L_{i-1}(x_i, x_{i-1})}
           {\gamma_{i-1}(x_{i-1}) K_i(x_{i-1}, x_i)}.
    \label{eq:smcwt}
\end{align}
The final key piece of information is to notice that, because the $L_i$ are
Markov kernels, $\pi_i$ is simply the marginal in $\vec{x_{i-1}}$ of
$\tilde{\pi}$. Therefore, a sample from $\tilde{\pi}_i$ automatically gets us a
sample from $\pi_i$, by considering only the $i$th component of $\vec{x_i}$.
These are all the ingredients we need to apply \gls{SIS}. The sequences of
kernels $L$ and $K$ should be chosen based on the problem at hand to minimize
the variance in the importance weights as well as possible. For a fixed choice
of $K_i$, the backward kernels $L_i$ which minimize this variance are called
the \defn{optimal} backward kernels. The full \gls{SMC} sampler algorithm is
presented as \cref{alg:smcsamp}. A resampling step is applied whenever the
\gls{ESS} of the population drops too low, as discussed in the previous
section.

@@

\glsreset{ABC}
{\color{blue}\uline{\acrlong{SMC}, and the \gls{SMC} sampler, were developed
for sampling from distributions which can be evaluated up to a normalizing
constant. TODO }

\uline{This section is organized as follows. First, we formalize the concept of a
mathematical model and discuss what it means to ``fit'' such models to data 
and how this is typically accomplished. Second, we introduce \gls{ABC}, which
was designed for situations like the one we face in this work where the most
commonly used means of fitting models cannot be applied. Finally, we describe
algorithms which can be used to implement \gls{ABC}, including a variant of
\gls{SMC}.}}

\subsection{Overview of ABC}
\label{subsec:abcoverview}

Most mathematical models are amenable to fitting via one or both of the
approaches, \gls{ML} or Bayesian inference, discussed above. However, there are
some, particularly in the domain of population
genetics~\autocite{beaumont2002approximate, beaumont2010approximate}, for which
calculation of either the likelihood or the product of the likelihood and the
prior may be infeasible. For example, one or both of these quantities may be
expressible only as an intractable integral. \Gls{ABC} is designed for such
cases, where standard likelihood-based techniques for model fitting cannot be
applied.

Ordinarily, Bayesian inference targets the posterior distribution $f(\theta
\mid y)$. That is, in the Bayesian framework, model parameters with higher
posterior density are ``better'' in the sense that they offer a more credible
explanation for the observed data. Approximate Bayesian computation offers an
alternative metric for parameter credibility, namely the similarity of
simulated datasets to the observed data. If datasets simulated under the model
closely resemble the real data, it follows that the model is a reasonable
approximation to the real-world process generating the observed data. More
formally, suppose we have a distance measure $\rho$ defined on the space of all
possible data our model could generate. \gls{ABC} aims to sample from the joint
posterior distribution of model parameters and simulated datasets $z$ which are
within some small distance $\varepsilon$ of the observed data $y$,
\[
  \pi_{\varepsilon}(\theta, z \mid y) =
  \frac{\pi(\theta) f(z \mid \theta) \I_{A_{\varepsilon, y}} (z)}
  {\int_{A_{\varepsilon, y} \times \Theta} \pi(\theta) f(z \mid \theta) \d \theta}.
\]
Here, $A_{\varepsilon, y}$ is an $\varepsilon$-ball around $y$ with
respect to $\rho$, $\Theta$ is the space of all possible model parameters, and
$\I$ is the indicator function~\autocite{marin2012approximate}. As we shall
see in the next section, this distribution can be sampled from exactly. The
word ``approximate'' derives from the assumption that, for a suitably chosen
distance $\rho$ and a small enough $\varepsilon$, the marginal in $z$ of
this distribution approximates the posterior of
interest~\autocite{marin2012approximate}. That is,
\[
  \int \pi_\varepsilon(\theta, z \mid y) \d z \approx f(\theta \mid y).
\]
The distribution $\pi_\varepsilon(\theta, z \mid y)$ is variously referred to
as the \textit{\gls{ABC} target distribution} or the \gls{ABC} approximation
to the posterior. The intuition for why this approximation might be reasonable
comes from the fact that, when $\varepsilon = 0$, the $\varepsilon$-ball around
$y$ should contain only $y$ itself, hence the integral on the left is exactly
equal to the posterior. Thus, by taking $\varepsilon$ small, we should attain
something close to the posterior {\color{blue}\uline{if $\rho$ captures the
similarity between datasets reasonably well. However, the accuracy of the
\gls{ABC} approximation depends heavily on the choice of distance function
\autocite{aeschbacher2012novel,blum2013comparative}.}}

@@

Rejection \gls{ABC} is easy to understand and implement, but it is not
generally computationally feasible. If the posterior is very different from the
prior, a very large number of samples may need to be taken in order to find a
simulated dataset which is close to $z$. The inefficiency is compounded
by the curse of dimensionality - the measure of the $\varepsilon$-ball around
$y$ decreases exponentially with the number of dimensions.
\gls{ABC}-\gls{MCMC} (\cref{alg:abcmcmc}) was designed to overcome these
hurdles~\autocite{marjoram2003markov}. The approach is similar to ordinary
Bayesian \gls{MCMC} (\cref{subsec:mfit}), except that a distance cutoff
replaces the likelihood ratio. That is, the transition probability between
states $x$ and $x'$ is defined as
\[
  \min\left(1, \frac{\pi(\theta') q(\theta', \theta)}
                    {\pi(\theta) q(\theta, \theta')} 
    \cdot \I_{A_{\varepsilon, y}}(z') \right).
\]

\begin{algorithm}
  \caption{\gls{ABC}-\gls{MCMC}.}
  \begin{algorithmic}
    \State Draw $\theta$ according to $\pi(\theta)$
    \Loop
      \State Propose $\theta'$ according to $q(\theta, \theta')$
      \State Simulate a dataset $\vec{z'}$ according to the model with
             parameters $\theta$
      \State Accept $\theta \gets \theta'$ with probability
      $\min \left( 1, 
       \dfrac{\pi(\theta') q(\theta', \theta)}
             {\pi(\theta\phantom{'}) q(\theta, \theta')} 
       \cdot \I_{A_{\varepsilon, y}}(z') \right)$
    \EndLoop
  \end{algorithmic}
  \label{alg:abcmcmc}
\end{algorithm}

@@

The most recently developed class of algorithm for \gls{ABC} is
\gls{ABC}-\gls{SMC}~\autocite{sisson2007sequential, beaumont2009adaptive}. As
with \gls{ABC}-\gls{MCMC}, the algorithm is a straightforward modification of
an existing Bayesian inference method, in this case the \gls{SMC} sampler
(\cref{subsec:smcsamp}). The sequence of target distributions is defined as
$\pi_i = \pi_{\varepsilon_i}(x \mid y)$ for a decreasing sequence of tolerances
$\varepsilon_i$. The intention is for the algorithm to progress smoothly
through a sequence of target distributions which ends at the \gls{ABC}
approximation to the posterior. As discussed in \cref{subsec:smcsamp}, the
choices of the kernels $K$ and $L$ is problem-specific, and so appropriate
kernels must be chosen for \gls{ABC}. Several options have been
proposed~\autocite{beaumont2009adaptive, sisson2007sequential,
del2012adaptive}.

>>>>

{\color{blue}\uline{
Recall that the primary objective of our work is to develop a statistical
inference method for estimating contact network parameters from transmission
trees. In particular, for a network model with parameters $\theta$ and an input
transmission tree $T$, we are interested in the posterior distribution}
\begin{align}
  \Pr(\theta \mid T) = \frac{\Pr(T \mid \theta) \Pr(\theta)}{\Pr(T)},
  \label{eq:post}
\end{align}
where we have adopted the common abuse of the symbol $\Pr$ to refer to a
probability density. \uline{As alluded to in} \cref{sec:obj}, \uline{both the
likelihood $\Pr(T \mid \theta)$ and the normalizing constant $\Pr(T)$ are
likely computationally intractable (this will be explained further in}
\cref{subsec:just}\uline{). Hence, rather than computing the posterior
distribution analytically, we will approximate it using a \defn{Monte Carlo}
approach. The fundamental idea behind Monte Carlo methods is succinctly
expressed by \textcite{liu2001theoretical}}:
\begin{quote}
    \uline{Monte Carlo's view of the world is that any probability distribution $\pi$, 
    regardless of its complexity, can always be \emph{represented} by a
    discrete sample from it. By ``represented'', we mean that any computation
    of expectations using $\pi$ can be replaced to an acceptable degree of
    accuracy by using the empirical distribution resulting from the discrete
    sample.}
\end{quote}
\uline{In other words, if we are able to sample enough points from a
distribution of interest, we will be able to make reasonably accurate
statements about the distribution itself. For example, the expected value of
the distribution can be estimated by the sample's population mean. The reason
Monte Carlo methods will be useful in this work is that algorithms exist for
obtaining samples from distributions that are analytically intractable and from
which direct sampling is not possible } (for a review see
\autocite{robert2004monte}). \uline{\Gls{SMC}}~\autocite{doucet2000sequential,
doucet2001introduction, liu2008monte} \uline{is one such algorithm.}

@@

\uline{In this work, the distribution of interest is the posterior distribution}
\ref{eq:post}. \uline{The particles are particular values of the parameters
$\theta$ of the contact network model being studied. If we were taking a
typical Bayesian Monte Carlo approach to this problem, the particles would end
up weighted by their posterior probability and distributed in such a way that
the weighted population was a reasonable representation of $\Pr(\theta \mid
T)$. In our case, due to the intractable likelihood, we will need to consider
an approximation to the posterior } (see \cref{sec:abc}). \uline{However, for
now, nothing is lost by assuming that our target distribution is the posterior
itself.}

\uline{The goal of this section of the introduction is to describe an algorithm
called the \gls{SMC} sampler~\autocite{del2006sequential}, which forms the
basis of the adaptive \gls{ABC}-\gls{SMC} algorithm we apply toward the main
objective of this work. We begin by describing \gls{SIS}, which is a precursor
to \gls{SMC} that samples from a sequence of distributions defined on spaces of
increasing dimension. We then describe \gls{SMC} itself, which extends
\gls{SIS} with a resampling step to fight particle degeneracy. Finally, we
outline the \gls{SMC} sampler, which allows \gls{SMC} to be applied to
sequences of distributions all defined on the same space. This terminology will
become clear as the methods are described. }}

{\color{red}\sout{Sequential Monte Carlo (SMC) is the name for a family of
statistical inference methods that rely on approximating probability
distributions of interest with large collections of \defn{particles}, here
denoted $\set{x^{(k)}}$}\autocite{doucet2000sequential, doucet2001introduction,
liu2008monte}. \sout{These collections or \defn{populations} are constructed to form
a \defn{Monte Carlo approximation} to some distribution of interest $\pi$,
meaning that the empirical distribution of the particles converges in
distribution to $\pi$ as the population size gets
large}\autocite{liu2001theoretical}. \sout{The word \defn{sequential} is used because
the particle populations are modified in an iterative fashion over time, for
example, to incorporate new evidence.}}

@@

Both algorithms involve progression through a sequence of related
distributions. For \gls{SIS}, these distributions are lower-dimensional
marginals of the target distribution, while for the \gls{SMC} sampler, they are
of the same dimension and constitute a smooth progression from an initial to a
final distribution. In both cases, the neighbouring distributions in the
sequence are related to each other in some way, and we can take advantage of
that relationship to create a sequence of importance distributions alongside
the sequence of targets. In \gls{SIS}, the neighbouring marginals
$\pi(\vec{x_i})$ and $\pi(\vec{x_{i+1}})$ were related by the conditional
density $\pi(x_i \mid \vec{x_{i-1}})$, which we used to inform the importance
distribution. In \gls{SMC}, the relationship between subsequent distributions
is less explicit, but it is assumed that they are related closely enough that
an importance distribution for $\pi_i$ can be easily transformed into one for
$\pi_{i+1}$. In particular, the sequence of importance distributions $\eta_i$
is constructed as
\begin{align}
  \label{eq:impint}
  \eta_i(x') = \int \eta_{i-1}(x) K_i(x, x') \d x,
\end{align}
where $K_i$ is a Markov kernel and the integral is over the space on which the
$\pi_i$ are defined. The choice of $K_i$ should be based on the perceived
relationship between $\pi_{i-1}$ and $\pi_i$. \textcite{del2006sequential}
propose the use of a \gls{MCMC} kernel with equilibrium distribution $\pi_i$.
That is,
\[
  K_i(x, x') = \max\left(1, \frac{q(x', x)\pi_i(x)}{q(x, x')\pi_i(x')}\right),
\]
where $q(x, x')$ is a proposal function such as a Gaussian distribution
centred at $x$ {\color{blue}\uline{from which $x'$ is drawn}} (see
\cref{chp:prelim}).

Although this method of building up $\eta$ appears straightforward, the
drawback is that the importance distribution itself becomes intractable. In
particular, evaluating $\eta_i(x)$ involves a $i$-dimensional integral of the
type in \cref{eq:impint}. As it is necessary to evaluate $\eta(x)$ pointwise to
perform \gls{IS}, this construction appears to have defeated the purpose of
providing an importance distribution for each $\pi_i$.
\textcite{del2006sequential} overcome this problem with two ``artificial''
objects. First, they propose the existence of \textit{backward} Markov kernels
$L_{i-1}(x_i, x_{i-1})$. For now, these kernels are arbitrary; they will later
be precisely defined on a problem-specific basis. Second, the authors define an
alternative sequence of target distributions
\[
  \tilde{\pi}_i(\vec{x_i}) = \pi_i(x_i) \prod_{k=1}^{i-1} L_k(x_{k+1}, x_k)
\]
of increasing dimension. This brings us back to the setting described above in
\cref{subsec:sis}, namely of building up an importance distribution of
dimension $d$ sequentially through lower-dimensional distributions. We can
write $\tilde{\pi}_i$ in terms of $\tilde{\pi}_{i-1}$ by noticing that
\begin{align*}
  \frac{\tilde{\pi}_i(\vec{x_i})}{\tilde{\pi}_{i-1}(\vec{x_{i-1}})} 
  = \frac{\pi_i(x_i) \prod_{k=1}^{i-1} L(x_{k+1}, x_k)}
  {\pi_{i-1}(x_{i-1}) \prod_{k=1}^{i-2} L(x_{k+1}, x_k)}
  = \frac{\pi_i(x_i) L(x_i, x_{i-1})}{\pi_{i-1}(x_{i-1})},
\end{align*}
and hence
\[
  \tilde{\pi}_i = \frac{\pi_i(x_i) L(x_i, x_{i-1})}{\pi_{i-1}(x_{i-1})} \cdot \tilde{\pi}_{i-1}.
\]
Therefore, the importance weights for these new targets are defined recursively as
\begin{align}
  w(\vec{x_i}) 
    &= \frac{\tilde{\pi}_i(\vec{x_i})}{\eta_i(\vec{x_i})} \\
    &= \frac{\tilde{\pi}_{i-1}(\vec{x_{i-1}}) \pi_i(x_i) L(x_i, x_{i-1})}
           {\eta_{i-1}(\vec{x_{i-1}}) \pi_{i-1}(x_{i-1}) K_i(x_{i-1}, x_i)} \\
    &= w(\vec{x_{i-1}}) \cdot
      \frac{\pi_i(x_i) L_{i-1}(x_i, x_{i-1})}
           {\pi_{i-1}(x_{i-1}) K_i(x_{i-1}, x_i)} \\
    &\propto w(\vec{x_{i-1}}) \cdot
      \frac{\gamma_i(x_i) L_{i-1}(x_i, x_{i-1})}
           {\gamma_{i-1}(x_{i-1}) K_i(x_{i-1}, x_i)}.
    \label{eq:smcwt}
\end{align}
The final key piece of information is to notice that, because the $L_i$ are
Markov kernels, $\pi_i$ is simply the marginal in $\vec{x_{i-1}}$ of
$\tilde{\pi}$. Therefore, a sample from $\tilde{\pi}_i$ automatically gets us a
sample from $\pi_i$, by considering only the $i$th component of $\vec{x_i}$.
{\color{blue}\uline{In fact, since the weight update} \cref{eq:smcwt} 
\uline{depends only on the $i$th and $i-1$st components of each particle, we do
not even need to keep track of the complete particles if we are only interested
in the final distribution.}} These are all the ingredients we need to apply
\gls{SIS}. The sequences of kernels $L$ and $K$ should be chosen based on the
problem at hand to minimize the variance in the importance weights as well
as possible. For a fixed choice of $K_i$, the backward kernels $L_i$ which
minimize this variance are called the \defn{optimal} backward kernels. The full
\gls{SMC} sampler algorithm is presented as \cref{alg:smcsamp}. A resampling
step is applied whenever the \gls{ESS} of the population drops too low, as
discussed in the previous section.

@@

\glsreset{ABC}

\subsection{Overview and motivation}
\label{subsec:abcoverview}

{\color{blue}\uline{\Acrlong{SMC}, and the \gls{SMC} sampler, were developed
for sampling from distributions which can be evaluated up to a normalizing
constant. We claim, and shall argue more thoroughly below}
(\cref{subsec:just}), \uline{that the posterior distribution}
\[
  \Pr(\theta \mid T) = \frac{\Pr(T \mid \theta) \Pr(\theta)}{\Pr(T)}
\]
\uline{for a contact network model with parameters $\theta$ and an input
transmission tree $T$ does not fall in this category (note that, as in the
previous section, we are using $\Pr$ to denote a probability density).
Therefore, \gls{SMC}, and other Bayesian and \gls{ML} techniques for fitting
mathematical models} (see \cref{chp:prelim}), \uline{cannot be directly
applied to our problem. In particular, \gls{MCMC} and the \gls{SMC} sampler are
designed for distributions $\pi$ which can be evaluated up to a normalizing
constant $Z$, that is, $\pi(x) = \gamma(x) / Z$. Both algorithms calculate the
ratio $\pi(x) / \pi(x') = \gamma(x) / \gamma(x')$ for a current value $x$ and
proposed updated value $x'$ - for \gls{MCMC}, this is part of the \acrlong{MH}
ratio, while for the \gls{SMC} sampler, it is required to calculate the
importance weights. In the context of Bayesian inference, this ratio is a
likelihood ratio, which must be calculated by computing the individual
likelihoods and dividing them. If the likeilhood is intractible, this is
clearly not a viable approach.}

\uline{\Gls{ABC}}~\autocite{rubin1984bayesianly, tavare1997inferring,
fu1997estimating}\uline{ was developed to estimate posterior distributions with
intractable likelihoods, which have arisen frequently in the domain of
population genetics} \autocite{beaumont2002approximate,
beaumont2010approximate}. \uline{\Gls{ABC} navigates around the intractable
likelihood by replacing the posterior as the target of inference by an
\emph{approximate} posterior. This distribution is constructed in such a way
that the ratios required for \gls{MCMC} and the \gls{SMC} sampler can be
computed. This conveniently allows us to apply the existing \gls{MCMC} and
\gls{SMC} algorithms with minimal changes. In the next section, we shall
demonstrate how this is done, but first we give the definition of the
approximate posterior. }}

{\color{red}\sout{Most mathematical models are amenable to fitting via one or
both of the approaches, \gls{ML} or Bayesian inference, discussed above.
However, there are some, particularly in the domain of population
genetics}~\autocite{beaumont2002approximate, beaumont2010approximate},
\sout{for which calculation of either the likelihood or the product of the
likelihood and the prior may be infeasible. For example, one or both of these
quantities may be expressible only as an intractable integral. \Gls{ABC} is
designed for such cases, where standard likelihood-based techniques for model
fitting cannot be applied.}}

{\color{red}\sout{Ordinarily, Bayesian inference targets the posterior
distribution $\Pr(\theta \mid y)$. That is, in the Bayesian framework,}} 
{\color{blue}\uline{By targeting the posterior distribution, Bayesian inference
makes the assertion that}} model parameters with higher posterior density are
``better'' in the sense that they offer a more credible explanation for the
observed data. The approximate posterior targeted by \gls{ABC} uses an
alternative metric for parameter credibility, namely the similarity of
simulated datasets to the observed data. If datasets simulated under the model
closely resemble the real data, it follows that the model is a reasonable
approximation to the real-world process generating the observed data. More
formally, {\color{blue}\uline{let $y$ be the observed data to which we are
trying to fit a model with parameters $\theta$. In the case of this work, the
data is a transmission tree $T$, but we shall stick with the generic variable
$y$ for now.}} Suppose we have a distance measure $\rho$ defined on the space
of all possible data our model could generate. \gls{ABC} aims to sample from
the joint posterior distribution of model parameters and simulated datasets $z$
which are within some small distance $\varepsilon$ of the observed data $y$,
\[
  \pi_{\varepsilon}(\theta, z \mid y) =
  \frac{\pi(\theta) \Pr(z \mid \theta) \I_{A_{\varepsilon, y}} (z)}
  {\int_{A_{\varepsilon, y} \times \Theta} \pi(\theta) \Pr(z \mid \theta) \d \theta}.
\]
Here, $A_{\varepsilon, y}$ is an $\varepsilon$-ball around $y$ with
respect to $\rho$, $\Theta$ is the space of all possible model parameters, and
$\I$ is the indicator function~\autocite{marin2012approximate}.
{\color{blue}\uline{The term $\Pr(z \mid \theta)$ appears to be the bothersome
likelihood again, but this will turn out not to be a problem because we are
simulating $z$ ourselves. In fact, it is possible to sample from this
distribution exactly (see next section).}}

{\color{blue}\uline{To return to the context of this thesis, the observed data
$y$ is an estimated transmission tree for a viral epidemic under investigation.
The model in question is a contact network model with parameters $\theta$. $z$
is a simulated transmission tree, obtained by first generating a contact
network under the model, and then simulating the spread of an epidemic over
that network. A transmission tree can be constructed by keeping track of who
infected whom during the simulated epidemic (further details will be given in
} \cref{subsec:nettree} \uline{). The distance function $\rho$ must compare two
transmission trees - the observed tree $y$, and the simulated tree $z$. We will
define this distance function using the tree kernel discussed in}
\cref{subsec:treeshape}. \uline{In words, the approximate posterior we consider
here is a distribution which assigns a joint probability density to model
parameters and simulated transmission trees under those parameters. The
probability density is proportional to the product of the prior on the
parameters, and the likelihood of the simulated transmission tree under those
parameters, but only if the simulated transmission tree is sufficiently close
to the true tree. Otherwise, the probability density is zero. }}

{\color{red}\sout{As we shall see in the next section, this distribution can
be sampled from exactly. The word ``approximate'' derives from the assumption
that,}}
for a suitably chosen distance $\rho$ and a small enough $\varepsilon$, the
marginal in $z$ of this distribution approximates the posterior of
interest~\autocite{marin2012approximate}. That is,
\[
  \int \pi_\varepsilon(\theta, z \mid y) \d z \approx f(\theta \mid y).
\]
The distribution $\pi_\varepsilon(\theta, z \mid y)$ is variously referred to
as the \textit{\gls{ABC} target distribution} or the \gls{ABC} approximation
to the posterior. The intuition for why this approximation might be reasonable
comes from the fact that, when $\varepsilon = 0$, the $\varepsilon$-ball around
$y$ should contain only $y$ itself, hence the integral on the left is exactly
equal to the posterior. Thus, by taking $\varepsilon$ small, we should attain
something close to the posterior {\color{blue}\uline{if $\rho$ captures the
similarity between datasets reasonably well. However, the accuracy of the
\gls{ABC} approximation depends heavily on the choice of distance function
\autocite{aeschbacher2012novel,blum2013comparative}.}}

@@

Rejection \gls{ABC} is easy to understand and implement, but it is not
generally computationally feasible. If the posterior is very different from the
prior, a very large number of samples may need to be taken in order to find a
simulated dataset which is close to $z$. The inefficiency is compounded
by the curse of dimensionality - the measure of the $\varepsilon$-ball around
$y$ decreases exponentially with the number of dimensions.
\gls{ABC}-\gls{MCMC} (\cref{alg:abcmcmc}) was designed to overcome these
hurdles~\autocite{marjoram2003markov}. The approach is similar to ordinary
Bayesian \gls{MCMC} (\cref{chp:prelim}), except that a distance cutoff
replaces the likelihood ratio. That is, the transition probability between
states $x$ and $x'$ is defined as
\[
  \min\left(1, \frac{\pi(\theta') q(\theta', \theta)}
                    {\pi(\theta) q(\theta, \theta')} 
    \cdot \I_{A_{\varepsilon, y}}(z') \right).
\]

\begin{algorithm}
  \caption{\gls{ABC}-\gls{MCMC}.}
  \begin{algorithmic}
    \State Draw $\theta$ according to $\pi(\theta)$
    \Loop
      \State Propose $\theta'$ according to $q(\theta, \theta')$
      \State Simulate a dataset $z'$ according to the model with parameters
             $\theta$
      \State Accept $\theta \gets \theta'$ with probability
      $\min \left( 1, 
       \dfrac{\pi(\theta') q(\theta', \theta)}
             {\pi(\theta\phantom{'}) q(\theta, \theta')} 
       \cdot \I_{A_{\varepsilon, y}}(z') \right)$
    \EndLoop
  \end{algorithmic}
  \label{alg:abcmcmc}
\end{algorithm}

@@

The most recently developed class of algorithm for \gls{ABC} is
\gls{ABC}-\gls{SMC}~\autocite{sisson2007sequential, beaumont2009adaptive}. As
with \gls{ABC}-\gls{MCMC}, the algorithm is a straightforward modification of
an existing Bayesian inference method, in this case the \gls{SMC} sampler
(\cref{subsec:smcsamp}). The sequence of target distributions is defined as
$\pi_i = \pi_{\varepsilon_i}(x \mid y)$ for a decreasing sequence of tolerances
$\varepsilon_i$. The intention is for the algorithm to progress smoothly
through a sequence of target distributions which ends at the \gls{ABC}
approximation to the posterior. {\color{blue}\uline{The initial importance
distribution, denoted as $\eta_0$ in} \cref{alg:smcsamp}, \uline{is taken to be}
$\Pr(\theta) \Pr(z \mid \theta)$. TODO}

As discussed in \cref{subsec:smcsamp}, the
choices of the kernels $K$ and $L$ is problem-specific, and so appropriate
kernels must be chosen for \gls{ABC}. Several options have been
proposed~\autocite{beaumont2009adaptive, sisson2007sequential,
del2012adaptive}.

\begin{algorithm}
  \caption{\gls{ABC}-\gls{SMC}.}
  \begin{algorithmic}
    \For {$k = 1$ to $n$}
      \State Sample $x_1^{(k)}$ from $\eta_1(x_1)$
      \Comment{Initialize the $k$th particle}
      \State $w^{(k)} \gets \gamma_1(x_1^{(k)}) / \eta_1(x_1^{(k)})$
      \State Normalize the weights so that $\sum w = 1$
    \EndFor
    \For {$i = 2$ to $d$}
      \For {$k = 1$ to $n$}
        \State Sample $x_i^{(k)}$ from $K(x_{i-1}^{(k)}, x_i)$
        \Comment Extend the $k$th particle
        \State $w^{(k)} \gets w^{(k)} \cdot \dfrac{\gamma_i(x_i) L_{i-1}(x_i, x_{i-1})}{\gamma_{i-1}(x_{i-1}) K_i(x_{i-1}, x_i)}$
      \EndFor
      \State Normalize the weights so that $\sum w = 1$
      \If {$\ESS(w) < T$}
        \Comment{$T$ is a user-defined threshold}
        \State Resample the particles according to $w$
        \For {$k = 1$ to $n$}
          \State $w^{(k)} \gets 1/n$
        \EndFor
      \EndIf
    \EndFor
  \end{algorithmic}
  \label{alg:abcsmc}
\end{algorithm}
